---
title: "karioCaS: Kraken Confidence Scores Exploration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{karioCaS: Kraken Confidence Scores Exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

***
## Introduction

Metagenomic classification using k-mer based tools like Kraken2 is incredibly fast, but it is notoriously prone to false-positive assignments. The standard approach of applying a single, global Confidence Score (CS) threshold across an entire sample is fundamentally flawed: **one size does not fit all**. Biological domains (Bacteria, Archaea, Viruses, and Eukaryota) have vastly different database representations, genome sizes, and k-mer uniqueness.

**karioCaS** (Kraken Confidence Scores for Reliable Domain-Specific Microbiota Inference) is an R package designed to solve this problem. It provides a robust analytical framework to evaluate taxonomic stability across multiple stringency levels, allowing researchers to find the exact inflection point where statistical noise is removed without losing true biological signal.

```{r setup}
library(karioCaS)
```

***

## 1. Setting up the Environment

karioCaS relies on a strict, automated directory architecture. To analyze your data, you must place your Kraken2/Metaphlan-style outputs inside a base directory, specifically within a folder named `000_mpa_original`.

Files must follow the naming convention: `SAMPLENAME_CSXX.mpa` (e.g., `SAMPLE01_CS09.mpa` for a Confidence Score of 0.9).

For this vignette, we will use the mock data bundled with the package, copying it to a temporary directory to comply with CRAN/Bioconductor file system policies.

```{r prep_data}
# 1. Create a temporary project directory
proj_dir <- file.path(tempdir(), "karioCaS_Tutorial")
dir.create(file.path(proj_dir, "000_mpa_original"), recursive = TRUE)

# 2. Locate and copy the bundled mock data
mock_data_path <- system.file("extdata/your_project_name/000_mpa_original", package = "karioCaS")
mock_files <- list.files(mock_data_path, full.names = TRUE)
file.copy(mock_files, file.path(proj_dir, "000_mpa_original"))

print(paste("Project Directory initialized at:", proj_dir))
```

***

## 2. Data Harmonization

The first step is to import the multiple stringency outputs into a cohesive Bioconductor `TreeSummarizedExperiment` (TSE) object. This step parses the taxonomic strings, infers the phylogenetic ranks, and standardizes the matrices.

```{r import_data}
# Import data and generate the TSE object (Step 000)
tse_object <- import_karioCaS(project_dir = proj_dir)

# The object is now ready for Bioconductor-native downstream analysis if desired
tse_object
```

*Note: This function automatically creates a `000_karioCaS_input_matrix` folder containing the `.rds` object and a human-readable `.tsv` audit trail.*

***

## 3. Visual Exploration (Steps 001 to 005)

karioCaS shines in its ability to generate publication-ready PDFs to visually inspect the behavior of taxa as stringency increases. All plotting functions automatically read from the harmonized data and save their outputs into dynamically generated subfolders.

```{r visual_exploration, eval=FALSE}
# Note: In an active R session, these commands will generate high-quality 
# PDF panels in your project directory.

# Evaluate how many taxa and reads "survive" as stringency increases
taxa_retention(project_dir = proj_dir)

# Evaluate taxonomic depth and Parent-to-Child resolution 
taxa_resolution(project_dir = proj_dir, parent_level = "Genus", child_level = "Species")

# Generate abundance heatmaps showing taxa extinction patterns at a given CS
heatmaps_karioCaS(project_dir = proj_dir, analysis_rank = "Species", confidence_score = 40)
```

***

## 4. Objective Thresholding: The Stability Index (SI)

While visual exploration is crucial, choosing the optimal Confidence Score manually introduces subjectivity. 

karioCaS implements a **Multi-Strategy Stability Index (SI)** engine. Instead of relying on rigid geometric formulas, it analyzes step-wise loss and regime shifts to mathematically identify the optimal threshold (Primary SI). It offers three distinct strategies:

* **Dynamic (Default):** Calculates the basal noise (attrition) of the sample's tail and dynamically identifies the threshold. Ideal for reducing false-positives (e.g. pathogen detection).
* **Segmented:** Uses a Broken-Stick regression model to find regime shifts. A balanced tolerance between false-positives and false-negatives (e.g. ecology investigations and discovering biological dark matter).
* **Manual:** Allows the user to define an acceptable step-wise loss toll per domain.

```{r optimize_cs}
# Calculate the Stability Index for all domains using the default Dynamic method
audit_df <- optimize_CS(project_dir = proj_dir, tax_level = "Species", method = "dynamic")

# View the mathematical assessment for the Primary SIs
head(audit_df[audit_df$SI_Type == "Primary_SI", c("Sample", "Domain", "CS", "Taxa_Count", "Step_Loss_Pct")])
```

The `optimize_CS()` function saves an objective audit trail (`.rds` and `.tsv`) and generates PDFs highlighting the mathematical shift on the decay curves.

***

## 5. The Ultimate Biological Mosaic

Once the optimal thresholds are defined mathematically, we extract the surviving taxa. Since Archaea might require a CS of 0.20 while Bacteria requires a CS of 0.40, `karioCaS` allows you to create a "Mosaic" `.mpa` file using specific thresholds for each domain.

By setting the arguments to `"auto"`, the package natively reads the `optimize_CS` audit file and applies the mathematically optimal *Primary SI* for each domain.

```{r retrieve_taxa}
# Generate the final mosaic using automated mathematical thresholds
retrieve_selected_taxa(
  project_dir = proj_dir,
  tax_level = "Species",
  CS_B = "auto",       # Will pull the Primary SI for Bacteria
  CS_A = "secondary",  # Attempts to pull a Secondary SI (or fallbacks to Primary)
  CS_E = 40,           # Manual override (strictly CS40)
  CS_V = 0             # Manual fallback (CS0)
)
```

The resulting `SAMPLENAME_karioCaS_Mosaic.mpa` file, saved in the `1000_final_selection` folder, is a highly refined, high-confidence taxonomic profile, scrubbed of statistical noise and ready for downstream ecological or clinical analysis.

```{r cleanup, include=FALSE}
# Clean up temporary directory
unlink(proj_dir, recursive = TRUE)
```

***

## Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```